Numerical modelling of the Cordilleran ice sheet

### 2014-08-09 kappa.tex
Adjust figure widths.
Rework and shorten abstract.

This doctoral thesis presents a study of the glacial history of northwestern
North America using numerical ice sheet modelling calibrated against field
evidence. The area is characterized by the steep topography of several mountain
ranges separated by large inter-montane depressions that were once covered by a
large-scale ice mass, the former Cordilleran ice sheet. Due to this complex
topographic setting, the Cordilleran ice sheet remains the least understood
Pleistocene ice sheet in terms of its extent, volume, and dynamics. Because of
the irregular topography on which the ice sheet formed, field studies have
often only had local or regional relevance. Numerical modelling, on the other
hand, allows quantitative reconstructions of whole ice sheet evolution based on
approximated physics of glacier flow.

This thesis presents simulations performed with PISM, a numerical ice sheet
model that approximates glacier flow by basal sliding and internal deformation,
and allows simulations of ice sheet growth and decay over multi-millenial time
scales. Sensitivity experiments have shown that simulations adopting climate
forcing from the North American Regional Reanalysis (NARR) yield ice sheet
extents that best fit the field constraints during the Last Glacial Maximum
(LGM). These simulations show that the geometry of the LGM Cordilleran ice
sheet can be largely explained by present-day temperature and precipitation
fields combined with palaeo-climate reconstructions if an accurate treatment of
seasonality, orographic precipitation, and daily temperature variability is
included in the ice sheet model.

In contrast to this short-lived, maximum ice sheet configuration, most of the
last glacial cycle is characterized by an intermediate state of glaciation
including isolated glaciers and ice caps covering the major mountain ranges of
the North American Cordillera. The largest and most persistent of these ice
sheet nucleation centres is located over the Skeena Mountains. Modelled
deglaciation conditions broadly match field evidence, and where they diverge,
such as over the Fraser Plateau, they indicate either a lack of important
processes in the model, or a more complex glaciation history than previously
thought. The model yields a readvance of the ice sheet during late glacial
times, which appears to match field evidence, before final retreat towards the
Skeena Mountains, which we identify as a key area to understanding glacial
dynamics of the Cordilleran ice sheet, highlighting the need for further
geological investigation of this region.

@@

\begin{figure}
  \includegraphics[width=80mm]{model-siassa}
  \caption{Combination of Shallow Ice Approximation (SIA) and Shallow Shelf
           Approximation (SSA) velocities in PISM, assuring a smooth transition
           between shear-dominant and longitudinal stress regimes. After
           \citet[Fig.~1]{Winkelmann.etal.2011}. Not to scale.}
  \label{fig:model-siassa}
\end{figure}

@@

\item The LGM Cordilleran ice sheet was a transient feature, out of balance
with contemporaneous atmospheric temperatures. Its maximum extent is the result
of a 50\,ka-long evolution. Peak glacial temperatures were lower than those
needed sustain a constant ice sheet geometry because the ice sheet growth was
finally inhibited by the deglacial transition to warmer temperatures .
Therefore, this maximum stage is not representative of average conditions, as
shown in the modelling by persistent ice caps throughout most of the last
glacial cycle.

>>>>

This doctoral dissertation presents a study of the glacial history of the North
American Cordillera using numerical ice sheet modelling calibrated against
field evidence. This area, characterized by the steep topography of several
mountain ranges separated by large inter-montane depressions, was once covered
by a large-scale ice mass: the former Cordilleran ice sheet. Because of the
irregular topography on which the ice sheet formed, geological studies have
often only had local or regional relevance, and thus the Cordilleran ice sheet
remains the least understood Pleistocene ice sheet in terms of its extent,
volume, and dynamics.

In this thesis, I present numerical simulations performed with the Parallel Ice
Sheet Model (PISM), that allow quantitative reconstructions of the former ice
sheet evolution based on approximated physics of glacier flow. These
simulations show that the geometry of the Last Glacial Maximum Cordilleran ice
sheet was largely controlled by sharp contrasts in regional temperature,
precipitation, and daily temperature variability associated with the presence
of the mountain ranges.

However, this maximum stage appear short-lived and out of balance with
contemporaneous climate. During most of the simulated last glacial cycle, the
North American Cordillera is characterized by an intermediate state of
glaciation including isolated glaciers and ice caps covering the major mountain
ranges, the largest of which is located over the Skeena Mountains. Numerically
modelled patterns of ice sheet growth and decay are diverse, and reveal a
higher complexity than that encapsulated in previous, conceptual models.

@@

\begin{figure}
  \includegraphics{model-siassa}
  \caption{Combination of Shallow Ice Approximation (SIA) and Shallow Shelf
           Approximation (SSA) velocities in PISM, assuring a smooth transition
           between shear-dominant and longitudinal stress regimes. After
           \citet[Fig.~1]{Winkelmann.etal.2011}. Not to scale.}
  \label{fig:model-siassa}
\end{figure}

@@

\item The LGM Cordilleran ice sheet was a transient feature, out of balance
with contemporaneous atmospheric temperatures. Its maximum extent is the result
of a 50\,ka-long evolution. Peak glacial temperatures were lower than those
needed sustain a constant ice sheet geometry because the ice sheet growth was
finally inhibited by the deglacial transition to warmer temperatures.
Therefore, this maximum stage is not representative of average conditions, as
shown in the modelling by persistent ice caps throughout most of the last
glacial cycle.
